Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-04-10 inc/photocathode/plasmon/plasmon.tex
merging comments from andreas on photocathode chapter

In the search for alternative mechanisms to reduce the intrinsic divergence of the electron beam upon photoemission, a possible area of interest was Plasmon-Assisted Photoemission (PAPE). 
As the name suggests, a plasmon is the quasiparticle of plasma oscillation.
When this occurs at the surface of a conducting material, the resultant quasiparticles are called ``surface plasmon-polaritons.''
Since these will be the only plasmons discussed in this thesis, the word ``plasmon'' will be used interchangeably for the name ``surface plasmon-polariton.''
Electrons in a conductor can couple with electromagnetic fields, in this case a laser field, and oscillate accordingly.
This overall oscillation of electrons near the surface of the conductor creates strong oscillatory local surface electric field \cite{cottam_introduction_2004,concepts_2002}.
In PAPE, the goal is use this field to both enhance the emission and `control' the divergence of photoemitted electrons through the combined action of photoemission barrier suppression and ponderomotive acceleration.
On the other hand, these oscillating electrons are nothing but an oscillating current, and thus if a large plasmon response is established, resistive heating is a major concern even for small the small resistivities typical of conductors.

This avenue of investigation was motivated by Zawadzka et al \cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors mount a thin gold foil on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the ``grating coupling'' geometry \cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Futher, in this front-side illuminated geometry the back surface is now free to be attached to a heat sink.

@@

Every conducting material has an intrinsic plasma frequency given by 
\begin{equation}
  \omega_{P} = \sqrt{\frac{\rho_{e} e^2}{\varepsilon_{0} m^*}} \,\text{,}
\end{equation}
where $\rho_{e}$ is the free-electron density and $m^*$ is the electron effective mass in the material.
And in the most simple (undamped) approximations, the dielectric function of the material may be written in terms of this constant
\begin{equation} \label{eq:dielectric_as_omega}
  \varepsilon(\omega) = 1 - \frac{ \omega_{P}^2 }{ \omega^2 } \,\text{.}
\end{equation}

@@

In the grating coupling scheme (shown in \ref{fig:plasmon_schematic}) a laser of wavelength $\lambda$ is incident on a material with a grating of period $d$, making an angle $\theta$ from the surface normal.
A surface plasmon is driven when the sum of the component of the laser wave-vector parallel to the material surface ($k_{Lx}$) and a fourier component ($k_{G} = 2 \pi / d$) of the grating matches the wave-vector of the surface plasmon
\begin{equation}
  k_{SP} = k_{Lx} + n k_{G} \,\text{,}
\end{equation}
where $n$ here is an integer.
This expression reduces to 
\begin{equation} \label{eq:plasmon_condition}
  \sqrt{ \frac{ \varepsilon }{ 1 + \varepsilon } } = \sin \theta + \frac{ n \lambda }{ d } \,\text{,}
\end{equation}
where $\lambda$ is the incident laser wavelength.
For a given grating period, \ref{eq:plasmon_condition} determines the angle of incidence that the laser must make with the surface in order to drive the plasmon.
Alternatively, it may be used to design the periodic surface structure, for example allowing for angles that the microscope column will accept, different driving orders ($n$) and even other Fourier components of the periodic surface structure.

@@

For proof-of-concept purposes the first experiment was run on a commercially available gold foil on glass holographic grating (750 lines/mm).
Like the Kretschmann geometry, a foil on glass grating is vulnerable to heating problems; a plasmon creates a large amount of heat but the glass is a poor heat conductor.
The aforementioned laser (emitting $\sim$200fs pulses of green light at a peak pulse intensity of 30MW/cm$^2$) was used to drive the plasmon.

@@

There are several conclusions that can be drawn from this experiment.
Most importantly, it shows that a plasmon was driven successfully given the angular dependence and large response, few other processes could have occurred.
The heat generated by the plasmon oscillation destroyed the grating, therefore it validated the decision not to pursue a Kretschmann geometry over fears of heat damage at the high heating rates of the laser-plasmon interaction.
Finally, it suggests that if using a thermally conductive and structurally sound photoemission material and mount for the plasmonic photocathode, PAPE might generate a large number of electrons (as evident by the bright flash).

@@

Collaborators at Argonne National Laboratory's Electron Microscopy Center, Material Science Division employed a defocused Gallium focused ion beam (FIB) (Zeiss 1540XB) to iteratively mill an approximately sinusoidal shape into the surface of a silicon wafer.
Several regions were created each 100x100$\mu$m in area, having a $\sim$1$\mu$m period and $\sim$10\% modulation depth.
By \ref{eq:plasmon_condition}, given a laser wavelength of 523nm ($\varepsilon_{Au} \approx -3.95$ \cite{johnson_optical_1972}), this choice of periodicity corresponds to an incidence angle of $\sim39^{\circ}$, which is within the acceptance range of our experimental setup (35-75$^{\circ}$) (see Section \ref{sec:gun_design}).
An SEM image of one such region is shown in \ref{fig:fib-si-sem}.
A 300nm film of gold was deposited onto the silicon using the Varian e-beam deposition chamber at the UIC Nano Core Facility (NCF); a thin layer of chromium was deposited between the silicon and gold as a binding agent.
As the gold film is several times thicker than the penetration depth of green light on gold, the laser radiation cannot interact with the silicon.
%TODO should I comment on the coating acting to smooth the surface to a proper sinusoid?

@@

As seen in \ref{fig:pape}, at low incident laser intensities, below a laser intensity of about 25MW/cm$^2$, the emission from the nano-patterned photocathode at the plasmon resonance condition is very similar to that observed from a flat gold surface neighboring the FIB-milled region.
Specifically, the strength of both emissions have a quadratic dependence on the incident laser intensity.
This is consistent with a two-photon field emission process --- it requires the absorption of two 2.37eV (green) photons for an electron to acquire sufficient energy to allow any significant field emission as gold has a work function of 5.1eV.
Moreover, both emissions have similar momentum distributions (\ref{fig:pape} A and E), again indicating a similar emission process.
The only difference is that the emission from the nano-patterned region is 3-4 times more efficient.
This is readily explained by the fact that the periodic surface modulation results in a local enhancement of the $\sim$1MV/m DC gun field at the peaks of the modulation which, through the Schottky effect, then presents a reduced barrier for field emission.

>>>>

In the search for alternative mechanisms to reduce the intrinsic divergence of the electron beam upon photoemission, a possible area of interest was Plasmon-Assisted Photoemission (PAPE). 
As the name suggests, a plasmon is the quasiparticle of plasma oscillation.
When this occurs at the surface of a conducting material, the resultant quasiparticles are called ``surface plasmon-polaritons.''
Since these will be the only plasmons discussed in this thesis, the word ``plasmon'' will be used interchangeably for the name ``surface plasmon-polariton.''
Electrons in a conductor can couple with electromagnetic fields, in this case a laser field, and oscillate accordingly.
This overall oscillation of electrons near the surface of the conductor creates strong oscillatory local surface electric field \cite{cottam_introduction_2004,concepts_2002}.
In PAPE, the goal is use this field to both enhance the emission and `control' the divergence of photoemitted electrons through the combined action of photoemission barrier suppression and ponderomotive acceleration.
On the other hand, these oscillating electrons are nothing but an oscillating current, and thus if a large plasmon response is established, resistive heating is a major concern even for the small resistivities typical of conductors.

This avenue of investigation was motivated by Zawadzka et al \cite{zawadzka_evanescent_2001} who indicate that when driving a surface plasmon on gold they witnessed a reduction in the emission angle of photoemitted electrons.
In their paper, the authors coat a thin gold film on a prism; the laser then ``back illuminates'' the gold through the prism, ejecting electrons on vacuum side of the foil.
This arrangement is called the Kretschmann geometry.
While this geometry is common, due to fears about the lack of heat conduction through the glass and the power needed to generate a sufficient electron beam current, we felt that for UEM applications we should explore the alternative ``grating coupling'' geometry \cite{kupersztych_ponderomotive_2001,kupersztych_anomalous_2005,li_surface_2013}.
In this geometry, the plasmon is driven on a periodic surface rather than a flat surface attached to glass.
Futher, in this front-side illuminated geometry the back surface is now free to be attached to a heat sink.

@@

Every conducting material has an intrinsic plasma frequency given by 
\begin{equation}
  \omega_{P} = \sqrt{\frac{\rho_{e} e^2}{\varepsilon_{0} m^*}} \,\text{,}
\end{equation}
where $\rho_{e}$ is the free-electron density and $m^*$ is the electron effective mass in the material.
And in the most simple (undamped) approximations, $\varepsilon$, the real part of the dielectric function of the material, may be written in the following form:
\begin{equation} \label{eq:dielectric_as_omega}
  \varepsilon(\omega) = 1 - \frac{ \omega_{P}^2 }{ \omega^2 } \,\text{.}
\end{equation}

@@

In the grating coupling scheme (shown in \ref{fig:plasmon_schematic}) a laser of wavelength $\lambda$ is incident on a material with a grating of period $d$, making an angle $\theta$ from the surface normal.
A surface plasmon is driven when the sum of the component of the laser wave-vector parallel to the material surface ($k_{Lx}$) and a Fourier component ($k_{G} = 2 \pi / d$) of the grating matches the wave-vector of the surface plasmon
\begin{equation}
  k_{SP} = k_{Lx} + n k_{G} \,\text{,}
\end{equation}
where $n$ here is an integer.
This expression reduces to 
\begin{equation} \label{eq:plasmon_condition}
  \sqrt{ \frac{ \varepsilon }{ 1 + \varepsilon } } = \sin \theta + \frac{ n \lambda }{ d } \,\text{,}
\end{equation}
where $\lambda$ is the incident laser wavelength.
For a given grating period, \ref{eq:plasmon_condition} determines the angle of incidence that the laser must make with the surface in order to drive the plasmon.
Alternatively, it may be used to design the periodic surface structure, for example allowing for angles that the microscope column will accept, different driving orders ($n$), and even other Fourier components of the periodic surface structure.

@@

For proof-of-concept purposes, the first experiment was performed on a commercially available gold foil on glass holographic grating (750 lines/mm).
Like the Kretschmann geometry, a foil on glass grating is vulnerable to heating problems; a plasmon creates a large amount of heat but the glass is a poor heat conductor.
The aforementioned laser (emitting $\sim$200fs pulses of green light at a peak pulse intensity of 30MW/cm$^2$) was used to drive the plasmon.

@@

There are several conclusions that can be drawn from this experiment.
Most importantly, it shows that a plasmon was driven successfully; given the angular dependence and large response, few other processes could have occurred.
The heat generated by the plasmon oscillation destroyed the grating, therefore it validated the decision not to pursue a Kretschmann geometry over fears of heat damage at the high heating rates of the laser-plasmon interaction.
Finally, it suggests that through the use of a thermally conductive and structurally sound photoemission material and mount for the plasmonic photocathode, PAPE might generate a large number of electrons (as evident by the bright flash).

@@

Collaborators at Argonne National Laboratory's Electron Microscopy Center, Material Science Division employed a defocused Gallium focused ion beam (FIB) (Zeiss 1540XB) to mill an approximately sinusoidal shape into the surface of a silicon wafer.
Several regions were created each 100x100$\mu$m in area, having a $\sim$1$\mu$m period and $\sim$10\% modulation depth.
By \ref{eq:plasmon_condition}, given a laser wavelength of 523nm ($\varepsilon_{Au} \approx -3.95$ \cite{johnson_optical_1972}), this choice of periodicity corresponds to an incidence angle of $\sim39^{\circ}$, which is within the acceptance range of our experimental setup (35-75$^{\circ}$) (see Section \ref{sec:gun_design}).
An SEM image of one such region is shown in \ref{fig:fib-si-sem}.
A 300nm film of gold was deposited onto the silicon using the Varian e-beam deposition chamber at the UIC Nano Core Facility (NCF); a thin layer of chromium was deposited between the silicon and gold as a binding agent.
As the gold film is many times thicker than the penetration depth of green light on gold and several times thicker than the penetration depth of the plasmon field, neither the laser radiation nor the plasmon field will couple with the silicon.
%TODO should I comment on the coating acting to smooth the surface to a proper sinusoid?

@@

As seen in \ref{fig:pape}, at low incident laser intensities, below a laser intensity of about 25MW/cm$^2$, the emission from the nano-patterned photocathode at the plasmon resonance condition is very similar to that observed from a flat gold surface neighboring the FIB-milled region.
Specifically, the strength of both emissions have a quadratic dependence on the incident laser intensity.
This is consistent with a two-photon field emission process --- it requires the absorption of two 2.37eV (green) photons for an electron to acquire sufficient energy to allow any significant field emission as gold has a work function of $\sim$5eV.
Moreover, both emissions have similar momentum distributions (\ref{fig:pape} A and E), again indicating a similar emission process.
The only difference is that the emission from the nano-patterned region is 3-4 times more efficient.
This is readily explained by the fact that the periodic surface modulation results in a local enhancement of the $\sim$1MV/m DC gun field at the peaks of the modulation which, through the Schottky effect, then presents a reduced barrier for field emission.

### 2013-04-10 inc/photocathode/setup/setup.tex
merging comments from andreas on photocathode chapter

To ensure that the transverse momentum distribution is observed, the YAG scintillator must be positioned at the back focal plane (the Fourier plane) of the lens system.
This is readily achieved by appropriate adjustment of the current $I$ in the magnetic lens coils; specifically, after determining the current required to focus the electron beam onto the scintillator, optical imaging relations together with the $I^2$ dependence of the magnetic lens strength are used to find the current that ensures that the average lens to scintillator distance equals the focal length of the lens system.
Simulation of the experimental apparatus (employing the extended AG model, see Section \ref{sec:external_forces} ), confirms that this procedure places the YAG scintillator at the back focal plane of the lens system to within a $\pm$0.5cm error (\ref{fig:transverse-measurement}).
This simulation also confirms that for our experimental conditions (less than 5,000 electrons/pulse and $\Delta x_0 > $ 30$\mu$m) space-charge effects do not influence the observations.
Additionally, and as expected from optical considerations, the modeling shows that the spatial beam size in the Fourier plane is independent of $\Delta x_0$, the rms laser spot size incident on the photocathode (i.e., the electron source size).

>>>>

To ensure that the transverse momentum distribution is monitored, the YAG scintillator must be positioned at the back focal plane (the Fourier plane) of the lens system.
This is readily achieved by appropriate adjustment of the current $I$ in the magnetic lens coils; specifically, after determining the current required to focus the electron beam onto the scintillator, optical imaging relations together with the $I^2$ dependence of the magnetic lens strength are used to find the current that ensures that the average lens to scintillator distance equals the focal length of the lens system.
Simulation of the experimental apparatus (employing the extended AG model, see Section \ref{sec:external_forces} ), confirms that this procedure places the YAG scintillator at the back focal plane of the lens system to within a $\pm$0.5cm error (\ref{fig:transverse-measurement}).
This simulation also confirms that for our experimental conditions (less than 5,000 electrons/pulse and $\Delta x_0 > $ 30$\mu$m) space-charge effects do not influence the observations.
Additionally, and as expected from optical considerations, the modeling shows that the spatial beam size in the Fourier plane is independent of $\Delta x_0$, the rms laser spot size incident on the photocathode (i.e., the electron source size).
All of the results in this chapter use the Fourier plane beam size as a measure of the initial rms transverse momentum $\Delta p_{\smallT}$.

### 2013-04-10 inc/photocathode/single_photon/single_photon.tex
merging comments from andreas on photocathode chapter

Single-photon photoemission is the most common photoemission process and the easiest to explain.
As described by Einstein in his famous paper on ``the photoelectric effect'' \cite{einstein_uber_1905}, as it is colloquially known, if an electron absorbs a photon whose energy is sufficient to promote the electron above the local vacuum level, it become free to leave the host material.
The material dependent minimum energy needed to eject an electron is termed its ``work function'' $\Phi$.
Any additional energy above the work function remains with the ejected electron, usually in the form of kinetic energy, and is called the ``excess photoemission energy'' $\Delta E$, for single photon photoemission, using a laser of frequency $\omega$ (photon energy $\hbar \omega$), this is simply,
\begin{equation}
  \Delta E = \hbar \omega - \Phi \,\text{.}
\end{equation}
The efficiency of photoemission $\eta_{{\scriptscriptstyle PE}}$, when $\Delta E$ is small compared to $\Phi$ may be written as,
\begin{equation}
  \eta_{{\scriptscriptstyle PE}} = A ( 1 - R ) \Delta E^2 \,\text{,}
\end{equation}
where $A$ is a material paramenter, and R is the photocathode surface reflectivity.

>>>>

Single-photon photoemission is the most common photoemission process and the easiest to explain.
As described by Einstein in his famous paper on ``the photoelectric effect'' \cite{einstein_uber_1905}, as it is colloquially known, if an electron absorbs a photon whose energy is sufficient to promote the electron above the local vacuum level, it becomes free to leave the host material.
The material dependent minimum energy needed to eject an electron is termed its ``work function'' $\Phi$.
Any additional energy above the work function remains with the ejected electron, usually in the form of kinetic energy, and is called the ``excess photoemission energy'' $\Delta E$; for single photon photoemission, using a laser of frequency $\omega$ (photon energy $\hbar \omega$), this is simply,
\begin{equation}
  \Delta E = \hbar \omega - \Phi \,\text{.}
\end{equation}
The efficiency of photoemission $\eta_{{\scriptscriptstyle PE}}$, when $\Delta E$ is small compared to $\Phi$ may be written as, %TODO ref
\begin{equation}
  \eta_{{\scriptscriptstyle PE}} = A ( 1 - R ) \Delta E^2 \,\text{,}
\end{equation}
where $A$ is a material paramenter (typically of the order of $\sim 10^{-4} \text{eV}^{-2}$ %TODO ref
), and $R$ is the photocathode surface reflectivity, which is itself a function of the polarization, frequency and angle of the incident light.
%TODO calculation of $\eta_{PE}$?

### 2013-04-10 inc/photocathode/two_photon_thermionic/two_photon_thermionic.tex
merging comments from andreas on photocathode chapter

Two-photon assisted thermionic emission (2PTE) occurs when two photons provide nearly enough energy to emit a cold electron, and therefore a hotter electron, in the tail of the Fermi distribution, may have enough energy to be emitted \cite{yen_thermally_1980}.
Of course this is even less efficient than two-photon photoemission, but as three-photon photoemission goes as $I^3$, 2PTE will still be the dominant photoemission mechanism.
For thermionic emission from electrons at a temperature $T_e$, the rms transverse momentum $\Delta p_{\smallT}$ is given by 
\begin{equation}
  \Delta p_{\smallT} = \sqrt{m k_B T_e} \,\text{,}
\end{equation}
where $k_B$ is the Boltzmann constant \cite{dowell_quantum_2009,jensen_emittance_2010}.
Since a high laser intensity is required to drive 2PTE, the laser itself will heat the electrons, therefore we expect the measured Fourier plane spot size to grow as the square root of the laser pulse energy.

>>>>

Two-photon assisted thermionic emission (2PTE) occurs when two photons provide nearly enough energy to emit a cold electron, and therefore a hotter electron, in the tail of the Fermi distribution, may have enough energy to be emitted \cite{yen_thermally_1980}.
Of course this is even less efficient than direct two-photon photoemission, but as three-photon photoemission goes as $I^3$, 2PTE will still be the dominant photoemission mechanism.
For thermionic emission from electrons at a temperature $T_e$, the rms transverse momentum $\Delta p_{\smallT}$ is given by 
\begin{equation}
  \Delta p_{\smallT} = \sqrt{m k_B T_e} \,\text{,}
\end{equation}
where $k_B$ is the Boltzmann constant \cite{dowell_quantum_2009,jensen_emittance_2010}.
Since a high laser intensity is required to drive 2PTE, the laser itself will heat the electrons, therefore, if the laser pulse duration is less than the thermalization time of the electron distribution, %TODO check this!
we expect the measured Fourier plane spot size to grow as incident laser pulse energy increases.
